\propertitle

### 2019-11-15 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-15T15:51:26+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep{gersteinetal1985} or
\enquote{interaction}
\citep{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep{bohteetal2000,amarietal2003} of neuronal
activity. In this section we discuss how our proposed
application bears on this kind of quantification. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

Let us translate this question into more precise terms first, because
\enquote{cooperativity}, \enquote{interaction}, and similar notions are
vague. A translation in terms of \emph{informational sufficiency} seems to
suit all these notions \parentext{for an intuitive understanding of
  sufficiency, which goes back to \cite{fisher1922}, see \eg\
  \cite[\sect~4.5]{bernardoetal1994}; \cite[\chap~8,
  \sect~14.2]{jaynes1994_r2003}; \cite{cifarellietal1982};
  \cite{kullbacketal1951}. For technical results and connections with the
  notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,cifarellietal1980,cifarellietal1981,diaconisetal1981,lauritzen1982_r1988,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,lauritzen2007,ayetal2015}}.
The idea is as follows. We want to assess our degree of belief about the
frequencies of the activities of the sample, or about the activity of the
sample in a new time bin. For this assessment we should use all measured
data we have. But it sometimes happens that dropping part of the data --
for example, the measured third- and higher-order moments -- leaves our
assessment almost unchanged: the dropped data are \emph{informationally
  irrelevant} or almost so. The remaining data -- for example, first and
second moments -- are then \emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

We can check how much a part of the data is irrelevant by simply dropping
it from the conditional of our probabilities, and see if the latter change
appreciably. When our probabilities are built with the maximum-entropy
method, this is done by dropping part of the constraints and see how the
resulting distribution changes. For example, we may only retain the first
two moments (means and correlations) or the first four moments. This
procedure is equivalent to calculating and comparing the probabilities of
our hypotheses about sufficiency, conditional the data, assuming their
pre-data probabilities to be equal. We perform this calculation below.

@@

In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from \citep[for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
The method introduced in the present work, however, reveals an issue with
this application at the sample level. The issue doesn't come from the
method itself but from a general fact of probability theory. If a
probability distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically this impossibility translates into a
system of $n$ independent equations in $m$ unknowns with $n>m$
\citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

In our example this means that if means and pairwise correlations or the
first four moments are informationally sufficient for a particular sample
from a brain area, then they \emph{cannot} be sufficient for the full
network of neurons in that area, or for a different sample. Vice versa, if
we assume that they are informationally sufficient for the full network,
then we must expect them \emph{not} to be sufficient for any recorded
sample thereof. Now, questions about sufficient statistics are meant to be
addressed to a whole brain area, not just to some specific, casually
recorded sample. Thus, paradoxically, maximum-entropy methods applied
\emph{at the sample level} give an answer opposite to the one we might
think they're giving.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-15T19:22:27+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been used as a way of
quantifying the \enquote{cooperativity} \citep[\eg][]{gersteinetal1985} or
\enquote{interaction}
\citep[\eg][]{martignonetal1995,schneidmanetal2006,shlensetal2006} or
\enquote{synchrony} \citep[\eg][we're only citing the first papers using
these terms]{bohteetal2000,amarietal2003} of neuronal
activity. In this section we discuss how our proposed
application bears on this kind of quantification. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our
% discussion holds for any sets of statistics.

Let us translate this question into more precise terms first, because
\enquote{cooperativity}, \enquote{interaction}, and similar notions are
vague. A translation in terms of \emph{informational sufficiency} seems to
suit all these notions \parentext{for the technical definition and
  intuitive understanding of sufficiency, which goes back to
  \cite{fisher1922}, see \eg\ \cite[\sect~4.5]{bernardoetal1994};
  \cite[\chap~8, \sect~14.2]{jaynes1994_r2003}; \cite{cifarellietal1982};
  \cite{kullbacketal1951}. For technical results and connections with the
  notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,cifarellietal1980,cifarellietal1981,diaconisetal1981,lauritzen1982_r1988,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,lauritzen2007,ayetal2015}}.
The idea is as follows. We want to assess our probabilities about the
frequencies of the activities of the sample, or about the activity of the
sample in a new time bin. For this assessment we should use all measured
data and statistics we have. But it can be the case that dropping part of
the data or statistics -- for example, the measured third- and higher-order
moments -- leaves our probabilities almost unchanged. This means that the dropped
statistics are \emph{informationally irrelevant} or almost so. The
remaining statistics -- for example, first and second moments -- are then
\emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

We can thus quantify how much a subset of statistics is relevant with
respect to a larger set by keeping only the subset in the conditional of
our probabilities, and checking how much the latter change. When our
probabilities are built with the maximum-entropy method, this means keeping
only part of the constraints (and re-maximizing the entropy). For example,
keeping only the first two moments (means and correlations) out of the
first four moments. This procedure is actually equivalent to comparing the
probabilities of our hypotheses about sufficiency, conditional the
\emph{full} data, assuming their pre-data probabilities to be equal. We
show this equivalence below and perform the calculations for our first data
set.

@@

Before applying this idea to neuronal activity, however, it is important to
keep in mind that \emph{informational sufficiency is not preserved under
  sampling}: if a probability distribution has some sufficient statistics,
then its marginals \emph{cannot} have the same sufficient statistics, and
vice versa, except for trivial cases such as uniform probability
distributions. This generally also holds among different marginals: if a
statistics is sufficient for some marginal, then it will not be sufficient
for another marginal. This impossibility is known in statistical mechanics:
marginals of Gibbs states are not Gibbs states \citep{maesetal1999}.
Mathematically this impossibility comes from the Pitman-Koopman theorem
\citep{koopman1936,pitman1936,darmois1935} connecting sufficient statistics
and the exponential family, and translates into the general impossibility
of solving a system of $n$ independent equations in $m$ unknowns with more
equations than unknowns, $n>m$ \citep[\sect~3.1]{portamanaetal2015}.

This fact is important in the analysis of neural activity when it is based
on small samples from brain regions. If means and pairwise correlations
seem informationally sufficient for a particular sample from a brain area,
then they may well be \emph{insufficient} for the full network of neurons
constituting that area. Vice versa, if they are informationally sufficient
for the full network, then they may turn out \emph{not} to be sufficient
(or less so) for any recorded sample thereof. Hence, if we wonder about
\enquote{cooperativity} or \enquote{interaction} of a brain area, we cannot
reliably use maximum-entropy distributions \emph{for a sample only}. Yet
the neuroscientific literature is besprinkled with such sample-level
analyses
\citep[\eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
